An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-07-20 simplified_sdc_intro/paper.tex
update Jacobian

To solve the reaction system implicitly, the ODE solver needs the Jacobian,
$\partial \Rb/\partial \Uc$.  We follow the method of \cite{castro_sdc}
and factor this into two pieces,
\begin{equation}
\label{eq:factored_jac}
{\bf J} = \frac{\partial \Rb}{\partial {\bf w }} \frac{\partial {\bf w}}{\partial \Uc}.
\end{equation}

Even though it has no reactive sources, we include $\rho$ in our conservative state
for the purposes of computing the Jacobian (we denote the conserved state used for the
Jacobian $\Uc^\prime$).
Writing this out for two species, $X_\alpha$ and $X_\beta$, we have
\begin{equation}
\Uc^\prime = \left ( \begin{array}{c} \rho \\ \rho X_\alpha \\ \rho X_\beta \\ \rho E \\ \rho e \end{array} \right )
\end{equation}
We take the intermediate state to be ${\bf w} = (\rho, X_\alpha, X_\beta,
K, e)^\intercal$, where $K$ is the kinetic energy:
\begin{equation}
K = \frac{1}{2} u^2
\end{equation}
Note: this intermediate state differs from what we used in \citet{castro_sdc} in that
we are using $e$ instead of $T$ as an independent variable now.  This follows from
the work shown in \citet{strang_rnaas} that explored integrating the Strang
reaction system with an energy equation.  If the network gives derivatives in terms
of temperature, we can convert as:
\begin{equation}
\left . \frac{\partial \phi}{\partial T} \right |_\rho =
  \left . \frac{\partial \phi}{\partial e} \right |_\rho 
  \left . \frac{\partial e}{\partial T} \right |_\rho =
  c_v \left . \frac{\partial \phi}{\partial e} \right |_\rho
\end{equation}
for some quantity $\phi(\rho, T)$.  Here we identified the specific heat
at constant volume, $c_v = \partial e/\partial T|_\rho$.

The Jacobian transformation $\partial \Uc^\prime/\partial {\bf w}$ is:
\begin{equation}
\frac{\partial \Uc}{\partial {\bf w}} = \left (
   \begin{array}{ccccc}
       1 & 0 & 0 & 0 & 0 \\
       X_\alpha & \rho & 0 & 0 & 0 \\
       X_\beta & 0 & \rho & 0 & 0  \\
       e + K &
                 \rho  e_{X_\alpha} & \rho e_{X_\beta} & \rho &
                 \rho \\
       e  &
                 \rho  e_{X_\alpha} & \rho e_{X_\beta} & 0 &
                 \rho
     \end{array}\right)
\end{equation}
where we use the following notation for compactness:
\begin{equation}
e_{X_k} = \dedXd
\end{equation}
We get the inverse
(computed via SymPy, see the included Jupyter notebook) as:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial {\bf w}}{\partial \Uc^\prime} = \left (
  \begin{array}{ccccc}
   1  & 0 & 0 & 0 & 0 \\
   - \frac{X_\alpha}{\rho} & \frac{1}{\rho} & 0 & 0 & 0 \\
   - \frac{X_\beta}{\rho} & 0 & \frac{1}{\rho} & 0 & 0 \\
   - \frac{K}{\rho} & 0 & 0 & \frac{1}{\rho} & -\frac{1}{\rho} \\
   \frac{\sum_k X_{k} e_{X_k}  - e}{ \rho} &
    -\frac{e_{X_\alpha}}{\rho} & -\frac{e_{X_\beta}}{\rho} & 0 & \frac{1}{\rho} \\
   \end{array}\right)
\renewcommand{\arraystretch}{1}
\end{equation}

The reaction vector is
\begin{equation}
\Rb(\Uc^\prime) = \left (  \begin{array}{c} 0 \\ \rho \omegadot_\alpha \\ \rho \omegadot_\beta \\ \rho \Sdot \\ \rho \Sdot \end{array} \right )
\end{equation}
and the Jacobian is computed as $\partial \Rb/\partial {\bf w}$:
\begin{equation}
\frac{\partial \Rb}{\partial {\bf w}} = \left (
  \begin{array}{ccccc}
     0 & 0 & 0 & 0 & 0 \\
     \omegadot_\alpha + \rho \frac{\partial \omegadot_\alpha}{\partial \rho} &
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\beta} & 0 & \rho \frac{\partial \omegadot_\alpha}{\partial e} \\
      %
     \omegadot_\beta + \rho \frac{\partial \omegadot_\beta}{\partial \rho} &
     \rho \frac{\partial \omegadot_\beta}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\beta}{\partial X_\beta} & 0 &  \rho \frac{\partial \omegadot_\beta}{\partial e} \\
     %
     \Sdot + \rho \frac{\partial \Sdot}{\partial \rho} & \rho \frac{\partial \Sdot}{\partial X_\alpha} & \rho \frac{\partial \Sdot}{\partial X_\beta} & 0 & \rho \frac{\partial \Sdot}{\partial e} \\
     %
     \Sdot + \rho \frac{\partial \Sdot}{\partial \rho} & \rho \frac{\partial \Sdot}{\partial X_\alpha} & \rho \frac{\partial \Sdot}{\partial X_\beta} & 0 & \rho \frac{\partial \Sdot}{\partial e} \\
  \end{array}
  \right )
\end{equation}

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via differencing, we found
that does not give reliable results.  Instead, we compute $\partial {\bf w}/\partial {\bf \Uc}^\prime$ analytically and $\partial {\bf R}/\partial {\bf w}$ numerically via one-sided differencing, following
the algorithm in \cite{lsode} to minimize numerical noise.

The final Jacobian is found by multiplying these two and removing the row corresponding to
$R(\rho)$ and the column corresponding to $\partial/\partial \rho$:
\begin{equation}
\frac{\partial \Rb}{\partial {\bf \Uc}} = \left (
  \begin{array}{ccccc}
    \frac{\partial \omegadot_\alpha}{\partial X_\alpha} -\frac{\partial \omegadot_\alpha}{\partial e} e_{X_\alpha} &
    \frac{\partial \omegadot_\alpha}{\partial X_\beta} -\frac{\partial \omegadot_\alpha}{\partial e}  e_{X_\beta} &
    0 &
    \frac{\partial \omegadot_\alpha}{\partial e} \\
     %
    \frac{\partial \omegadot_\beta}{\partial X_\alpha} -\frac{\partial \omegadot_\beta}{\partial e} e_{X_\alpha} &
    \frac{\partial \omegadot_\beta}{\partial X_\beta} -\frac{\partial \omegadot_\beta}{\partial e}  e_{X_\beta} &
    0 &
    \frac{\partial \omegadot_\beta}{\partial e} \\
     %
     \frac{\partial \Sdot}{\partial X_\alpha} - \frac{\partial \Sdot}{\partial e} e_{X_\alpha} &
     \frac{\partial \Sdot}{\partial X_\beta} - \frac{\partial \Sdot}{\partial e} e_{X_\beta} &
     0 &
     \frac{\partial \Sdot}{\partial e} \\
     %
     \frac{\partial \Sdot}{\partial X_\alpha} - \frac{\partial \Sdot}{\partial e} e_{X_\alpha} &
     \frac{\partial \Sdot}{\partial X_\beta} - \frac{\partial \Sdot}{\partial e} e_{X_\beta} &
     0 &
     \frac{\partial \Sdot}{\partial e} \\
  \end{array}
  \right )
\end{equation}

>>>>

To solve the reaction system implicitly, the ODE solver needs the
Jacobian, $\partial \Rb/\partial \Uc^\prime$, where $\Uc^\prime =
(\rho X_k, \rho e)^\intercal$ is the subset of the conserved variables we are
integrating.  We follow the method of \cite{castro_sdc} and factor
this into two pieces,
\begin{equation}
\label{eq:factored_jac}
{\bf J} = \frac{\partial \Rb}{\partial {\bf w }} \frac{\partial {\bf w}}{\partial \Uc^\prime}.
\end{equation}
Writing this out for two species, $X_\alpha$ and $X_\beta$, we have
\begin{equation}
\Uc^\prime = \left ( \begin{array}{c} \rho X_\alpha \\ \rho X_\beta \\ \rho e \end{array} \right )
\end{equation}
For interfacing with the reaction network, we use a simpler set of variables,
\begin{equation}
{\bf w} = \left ( \begin{array}{c} X_\alpha \\  X_\beta \\ e \end{array} \right )
\end{equation}
Note: this intermediate state differs from what we used in \citet{castro_sdc} but agrees with that from \cite{strang_rnaas} in that
we use $e$ instead of $T$ as the independent variable.  If the network gives derivatives in terms
of temperature, we can convert as:
\begin{equation}
\left . \frac{\partial \phi}{\partial T} \right |_\rho =
  \left . \frac{\partial \phi}{\partial e} \right |_\rho 
  \left . \frac{\partial e}{\partial T} \right |_\rho =
  c_v \left . \frac{\partial \phi}{\partial e} \right |_\rho
\end{equation}
for some quantity $\phi(\rho, T)$.  Here we identified the specific heat
at constant volume, $c_v = \partial e/\partial T|_\rho$.

The Jacobian transformation $\partial \Uc^\prime/\partial {\bf w}$ is:
\begin{equation}
\frac{\partial \Uc^\prime}{\partial {\bf w}} = \left (
   \begin{array}{ccc}
       \rho & 0 & 0 \\
       0 & \rho & 0  \\
       \rho  e_{X_\alpha} & \rho e_{X_\beta} & \rho \\
     \end{array}\right)
\end{equation}
where we use the following notation for compactness:
\begin{equation}
e_{X_k} = \dedXd
\end{equation}
We get the inverse
(computed via SymPy, see the included Jupyter notebook) as:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial {\bf w}}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
   \frac{1}{\rho} & 0 & 0 \\
   0 & \frac{1}{\rho} & 0 \\
    -\frac{e_{X_\alpha}}{\rho} & -\frac{e_{X_\beta}}{\rho} & \frac{1}{\rho} \\
   \end{array}\right)
\renewcommand{\arraystretch}{1}
\end{equation}

The reaction vector is
\begin{equation}
\Rb(\Uc^\prime) = \left (  \begin{array}{c} \rho \omegadot_\alpha \\ \rho \omegadot_\beta \\ \rho \Sdot \end{array} \right )
\end{equation}
We take all the quantities to be functions of $\rho$, $e$, and $X_k$,
but since $\rho$ doesn't change from the reactions (the reactive
source of the continuity equation is zero), it is held constant in
these derivatives.  The Jacobian is computed as $\partial \Rb/\partial
{\bf w}$:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial {\bf w}} = \left (
  \begin{array}{ccc}
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\beta} & 
     \rho \frac{\partial \omegadot_\alpha}{\partial e} \\
      %
     \rho \frac{\partial \omegadot_\beta}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\beta}{\partial X_\beta} &
     \rho \frac{\partial \omegadot_\beta}{\partial e} \\
     %
     \rho \frac{\partial \Sdot}{\partial X_\alpha} &
     \rho \frac{\partial \Sdot}{\partial X_\beta} &
     \rho \frac{\partial \Sdot}{\partial e} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via differencing, we found
that does not give reliable results.  Instead, we compute $\partial {\bf w}/\partial {\bf \Uc}^\prime$ analytically and $\partial {\bf R}/\partial {\bf w}$ numerically via one-sided differencing, following
the algorithm in \cite{lsode} to minimize numerical noise.
The final Jacobian is found by multiplying these two:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
    \frac{\partial \omegadot_\alpha}{\partial X_\alpha} -\frac{\partial \omegadot_\alpha}{\partial e} e_{X_\alpha} &
    \frac{\partial \omegadot_\alpha}{\partial X_\beta} -\frac{\partial \omegadot_\alpha}{\partial e}  e_{X_\beta} &
    \frac{\partial \omegadot_\alpha}{\partial e} \\
     %
    \frac{\partial \omegadot_\beta}{\partial X_\alpha} -\frac{\partial \omegadot_\beta}{\partial e} e_{X_\alpha} &
    \frac{\partial \omegadot_\beta}{\partial X_\beta} -\frac{\partial \omegadot_\beta}{\partial e}  e_{X_\beta} &
    \frac{\partial \omegadot_\beta}{\partial e} \\
     %
     \frac{\partial \Sdot}{\partial X_\alpha} - \frac{\partial \Sdot}{\partial e} e_{X_\alpha} &
     \frac{\partial \Sdot}{\partial X_\beta} - \frac{\partial \Sdot}{\partial e} e_{X_\beta} &
     \frac{\partial \Sdot}{\partial e} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}
